---
title: "Assigment of mutation signatures following HDP component extraction"
author: "Chiraag Kapadia, Daniel Leongamornlert, Nick Williams"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE)
knitr::opts_knit$set(root.dir = '~/Desktop/somatic_mouse/scripts/')
```

### Summary

0.  Import HDP output for analysis
1.  Assign identities to HDP clusters
2.  Reattribute mutations on the unassignent components and assess fit
3.  Assignment extracted signatures to tree branches
4.  Aggregate signature-specific mutation burden over life
5.  Branch location signature accrual over time

#### Library and support script dependencies

```{r message=FALSE, warning=FALSE}
library(sigfit)
library(MutationalPatterns)
library(scales)
library(dplyr)
SPECIES_BUILD="mm10"
SPECIES="Mmusculus"
source("revised_sigs.R")
```

#### Import HDP output for analysis

```{r warning=FALSE}
setwd("~/Desktop/HDP/")
mc.HSC.MPP_condensedallNOCOLON.hSorted_tissue_prior_multi_extract= readRDS("mc.HSC.MPP_condensedallNOCOLON.hSorted_tissue_prior_multi_extract.rds")
tissuemut_count.raw=readRDS("tissuemut_count.raw.rds") 
```

### 1. Assign identities to HDP clusters

```{r}
pcawg=get_pcawg60()

hdpinfo=extract_details_from_hdpres(mc.HSC.MPP_condensedallNOCOLON.hSorted_tissue_prior_multi_extract,
                                    priorsigs = pcawg, #any  
                                    mutcount = tissuemut_count.raw,
                                    threshold = .95) # cosime sim of 0.95 to match a known signature
plot_96_profile(hdpinfo$signatures)
hdpinfo$csm_max

```

We note that N1 contains C>A mutations resembling SBS18, and N2 constains T>G mutations and a profile resembling SBS5 mixed with SBS1. 

### 2. Reattribute mutations on the unassignent components and assess fit

```{r message=FALSE, warning=FALSE}
sfres = fit_signatures(t(round(hdpinfo$signatures*1e6)),
                       t(pcawg[,c("SBS1","SBS18","SBS5")]), #Matched priors
                       iter=10000)
reconstructions =retrieve_pars(sfres, "reconstructions")

plot_compare_profiles(rowSums(round(hdpinfo$signatures*1e6)),
                      rowSums(t(reconstructions$mean)),
                      profile_names = c("Original","Sigfit\nReconstructed") )
```

### 3. Assign mutations to tree branches, plot trees

Supporting scripts and tree objects
```{r message=FALSE, warning=FALSE}
source("treebased_signature_attribution.R")
source("load_and_annotate_tree.R")
source("trees.R")
library(cowplot)
library(phangorn)
library(lme4)
library(ggeffects)
pcawg=get_pcawg60()
tree_list = readRDS("~/Desktop/somatic_mouse/tree_data/all_tree_list.rds")
```

```{r, message=FALSE, warning=FALSE, echo=FALSE,results='hide',fig.keep='all'}
samples_hsc <- c("MD7180","MD7181","MD7182","MD7183","MD7184","MD7185")
PDD=tree_list[samples_hsc]
lapply(PDD, function(p){ 
  p$pdx$dat$info=get_clade_sigs_mouse(p,prior=as.matrix(pcawg[, c("SBS1","SBS5","SBS18")]),extracats=c())
  ww=plot_mouse_sig_tree_posterior(p,prior=as.matrix(pcawg[, c("SBS1","SBS5","SBS18")]),                                   focal.signature = "SBS18",b.ggbar_separate = T,lwd=1.3,default_edge_color="grey30")
})
```

### 4. Aggregate signature changes over life

```{r, eval=F, echo=T}
PDD2=lapply(PDD,function(PD){
  PD=make_ultrametric_PD_chiraag(PD) 
  PD$pdx$dat$info=get_clade_sigs(PD,prior=pcawg[,c("SBS1","SBS5","SBS18")],extracats = c());
  PD=add_v2_sig_to_pdx(PD,prior=pcawg[,c("SBS1","SBS5","SBS18")],info = PD$pdx$dat$info);PD
  })
```

```{r}
PDD2=readRDS("~/Desktop/somatic_mouse/data/signature_annotated_trees.rds")
df=do.call("rbind",lapply(PDD2,function(PD) get_sig_burdens(PD,priors = pcawg[,c("SBS1","SBS5","SBS18")])$df))

#Wrangle for ggplot..
#Whereever age is added, some are off..
df$age_at_sample_exact[which(grepl("80|81|82",df$patient))]<-30
df$age_at_sample_exact[which(grepl("83|84|85",df$patient))]<-3
df=subset(df, select = -c(nsub_adj) )
df_subset=df[which(!grepl("MPP",df$patient)),];comp="HSC"
# df_subset=df[which(grepl("MPP",df$patient)),];comp="MPP"

df_pivot = df_subset %>% pivot_longer(names_to="burden_type",values_to = "burden",
                           cols = c(sprintf("nsub_adj_%s",c("SBS1","SBS5","SBS18"))))

list_fixed=list()
pr=do.call("rbind",
    lapply(c(sprintf("nsub_adj_%s",c("SBS1","SBS5","SBS18"))), function(sbs){
  dfff=df_pivot[which(df_pivot$burden_type==sbs),]
  lmer1 =  lmer(burden ~ age_at_sample_exact + (0 + age_at_sample_exact|donor), data = dfff)
  print(summary(lmer1))
  pr=ggpredict(lmer1, "age_at_sample_exact")
  colnames(pr)[1]<-"age_at_sample_exact"
  lm_fixef=as.data.frame(fixef(lmer1))
  list_fixed[[sbs]]<<-data.frame(formula=sprintf("burden ~ %s*age + %s",round(lm_fixef[[1]][2],2),round(lm_fixef[[1]][1],2)),
                                 burden_type=sbs )
  pr$burden_type=sbs
  pr
}))
list_fixed=do.call("rbind",list_fixed)

list_fixed$y=0;list_fixed$y<-c(max(df_pivot$burden)-10,max(df_pivot$burden)-30,max(df_pivot$burden)-50)

colorscheme_nsub=data.frame(COL=(c("grey30", "#BB5566","#004488","#DDAA33")),SIG=(c("nsub_adj","nsub_adj_SBS1","nsub_adj_SBS5","nsub_adj_SBS18")))
ggplot(df_pivot,aes(x=age_at_sample_exact,y=burden,col=burden_type))+
  # geom_violin(position = position_dodge(width = 5))+
  # geom_point()+
  # geom_violin(aes(col = burden_type,group=age_at_sample_exact), position = position_dodge(width = 5))+
  geom_point(position =  position_jitterdodge(dodge.width = 2,jitter.width=.5, seed = 1))+
  ylab("SNVs")+ggtitle(sprintf("%s Sig burden",comp))+
  xlab("Age (months)")+
  # geom_smooth(method="lm",fullrange=T)+
  geom_line(data =pr, aes(y = predicted,col=burden_type),size=1) +
  geom_ribbon(data=pr, aes(y = NULL, ymin = conf.low, ymax = conf.high,
                   color = burden_type, fill = NULL
  ),alpha = .15)+
  geom_text(data=list_fixed,aes(x=3,y=y,col=burden_type,label=formula),hjust=0)+
  theme_minimal_hgrid()+
  # stat_regline_equation(
  #   aes(label =  paste(..eq.label.., ..adj.rr.label.., sep = "~~~~")),
  #   formula = formula
  # )+  
  scale_color_manual(breaks=colorscheme_nsub$SIG, values=colorscheme_nsub$COL)
```

### 5. Branch location signature accrual over time

```{r}
PDD2=readRDS("~/Desktop/somatic_mouse/data/signature_ultrametric_trees.rds")
df2=do.call("rbind",lapply(PDD2,function(PD) get_sig_burdens(PD,priors = pcawg[,c("SBS1","SBS5","SBS18")])$traj %>% mutate(donor=PD$patient)))
df2$age_at_sample_exact<-0
df2$age_at_sample_exact[which(grepl("80|81|82",df2$donor))]<-30
df2$age_at_sample_exact[which(grepl("83|84|85",df2$donor))]<-3

df2_subset=df2[which(!grepl("MPP",df2$donor)),];comp="HSC"
# df2_subset=df2[which(grepl("MPP",df2$donor)),];comp="MPP"

df2_subset_old=df2_subset[which(df2_subset$age_at_sample_exact==30),]; #comp="Aged"
df2_subset_young=df2_subset[which(df2_subset$age_at_sample_exact==3),]; #comp="Young"

for(age in c("young","old")){
  for(whereinbranch in c("start","mid","end")){
    colorscheme_nsub=data.frame(COL=(c("#BB5566","#004488","#DDAA33")),SIG=sprintf("%s_%s",c("SBS1","SBS5","SBS18"),whereinbranch))
    
    dff=df2_subset #Do it with all ages.
    dff=dff[which(dff$SBS5_end!=0),]
    
    dff_tips_aged=dff[which(round(dff$end,1)==135.0),] %>% pivot_longer(names_to="burden_type",values_to = "interim_burden",cols = sprintf("%s_%s",c("SBS1","SBS5","SBS18"),whereinbranch))
    dff_tips_young=dff[which(round(dff$end,1)==15.0),] %>% pivot_longer(names_to="burden_type",values_to = "interim_burden",cols = sprintf("%s_%s",c("SBS1","SBS5","SBS18"),whereinbranch))
    dff_tips=dff[which(round(dff$end,1)==15.0 | round(dff$end,1)==135.0),] %>% pivot_longer(names_to="burden_type",values_to = "interim_burden",cols = sprintf("%s_%s",c("SBS1","SBS5","SBS18"),whereinbranch))
    
   p=ggplot(dff %>% pivot_longer(names_to="burden_type",values_to = "interim_burden",cols = sprintf("%s_%s",c("SBS1","SBS5","SBS18"),whereinbranch)),
           aes(x=end,y=interim_burden,col=burden_type))+
      geom_point(position =  position_jitterdodge(dodge.width = 0,jitter.width=1, seed = 1),size=0.5)+
     scale_x_continuous(breaks = seq(0,130,26))+
      ylab(sprintf("Branch %s burden",whereinbranch))+
      theme_minimal_grid()+
     theme(legend.position = "none")+
      ggtitle(sprintf("%s branches, burden at branch %s, %s",comp,whereinbranch,age))+
      scale_color_manual(breaks=colorscheme_nsub$SIG, values=colorscheme_nsub$COL)
    plot(p)
  }
}

```


