---
title: "Polytomy analysis and visualization"
author: "Nick Williams, minor changes by Chiraag Kapadia"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE)
knitr::opts_knit$set(root.dir = './scripts/')
```

### Summary

0.  Define functions.
1.  Generate top-of-tree plots, as shown in Extended Data Fig. 3

#### Library and support script dependencies

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(cowplot)
library(kableExtra)
source("load_and_annotate_tree.R")
source("trees.R")
source("plotting_utils.R")

```

#### Import tree objects for analysis, intialize necessary variables.

```{r}
tree_list = readRDS("~/Desktop/somatic_mouse/tree_data/all_tree_list.rds")
```

### 0. Define functions

```{r warning=FALSE}
analyse_polytomies=function(PD,nmuts=10){
  tree=PD$pdx$tree_ml
  tree$edge.length=tree$el.snv
  tree=drop.tip(tree,"zeros")
  nh=nodeHeights(tree)
  NL=length(which(nh[,1]<nmuts & nh[,2]>=nmuts))
  ##We count how many descendents an edge has and then how many of each polytomy class
  polytomy_table=as.data.frame(table(table(tree$edge[which(nh[,1]<=nmuts),1])))
  colnames(polytomy_table)=c("polytomy_number","count")
  pt=polytomy_table
  pt$count=as.integer(pt$count)
  pt$polytomy_number=as.integer(as.character(pt$polytomy_number))
  pt$ndivisions=pt$polytomy_number-1
  pt$nzero=pt$polytomy_number-2
  pt$nbranches=pt$ndivisions*2
  totalbranches=sum(pt$nbranches*pt$count)
  totalzerobranches=sum(pt$nzero*pt$count)
  rateinf=infer_cd_rate(totalbranches,totalzerobranches)
  pt$patient=PD$patient
  list(pt=pt,res=cbind(data.frame(patient=PD$patient,number_of_lineages=NL,RATE_UB=nmuts/log2(NL)),rateinf))
}

infer_cd_rate=function(totalbranches,totalzerobranches){
  rate=-log(totalzerobranches/totalbranches)
  res=binom.test(totalzerobranches,totalbranches,conf.level = 0.99)
  rate_lb=-log(res$conf.int[2])
  rate_ub=-log(res$conf.int[1])
  data.frame(totalbranches=totalbranches,totalzerobranches=totalzerobranches,rate=rate,rate_lb=rate_lb,rate_ub=rate_ub)
}

get_tree_top=function(tree,height){
  nh=nodeHeights(tree)
  root=length(tree$tip.label)+1
  newedge=tree$edge[which(nh[,1]<=height),]
  nh=nh[which(nh[,1]<=height),]
  edge.length=ifelse(nh[,2]>height,height,nh[,2])-nh[,1]
  newtips=which(!(newedge[,2] %in% newedge[,1]))
  newroot=length(newtips)+1
  newedgeo=newedge
  ##Start at the root and renumber edges..
  k=1
  p=newroot
  nnodes=0
  depth=0
  dat=list(newedge=newedge,nnodes=nnodes,p=p,depth=depth,k=k)
  update_newedge=function(node,dat){
    if(dat$depth>800){
      stop("recursion depth exceeded")
    }
    dat$depth=dat$depth+1
    kids=which(newedgeo[,1]==node)
    N=length(kids)
    idx=which(newedgeo[,2]==node)
    if(N==0){
      if(length(idx)!=1){
        stop("Badly formed input tree")
      }
      dat$newedge[idx,2]=dat$k
      dat$k=dat$k+1
    }else{
      if(length(idx)>0){
        dat$newedge[idx,2]=dat$p
      }
      dat$nnodes=dat$nnodes+N
      dat$newedge[kids,1]=dat$p
      dat$p=dat$p+1
      for(i in kids){
        dat=update_newedge(newedgeo[i,2],dat)
      }
    }
    dat
  }
  dat=update_newedge(root,dat)
  out=list(edge=dat$newedge,Nnode=as.integer(dat$nnodes-dat$k+2),edge.length=edge.length,tip.label=sprintf("t%d",1:(dat$k-1)))
  mode(out$edge)="integer"
  class(out)="phylo"
  ##checkValidPhylo(out)
  out
}
```

### 1. Assess polytomy counts

The identified polytomies across all patients are as follows:

```{r message=FALSE, warning=FALSE}
PDD=tree_list[c("MD7180","MD7181","MD7183", "MD7182","MD7184","MD7185")] 
polytomies=do.call("rbind",lapply(PDD,function(x) analyse_polytomies(x,nmuts = 12)$pt))
kable(polytomies, "html") %>% kable_styling("striped") %>% scroll_box(width="95%")
```

Restricting to mice with enough early branches (10) the per mice, inferred minimum and maximum rates are as follows:

```{r message=FALSE, warning=FALSE}
polyres=do.call("rbind",lapply(PDD,function(x) analyse_polytomies(x)$res)) %>% filter(totalbranches>10)
kable(polyres, "html") %>% kable_styling("striped") %>% scroll_box(width="95%")
```

Pooling the results for the lower bounds across all 5 mice we get:

```{r message=FALSE, warning=FALSE}
kable(infer_cd_rate(sum(polyres$totalbranches),sum(polyres$totalzerobranches)) ,"html") %>% kable_styling("striped") %>% scroll_box(width="95%")

```

### 2. Plot the trees and the polytomies
```{r}
par(mfrow=c(5,2),oma=c(2,1,1,1))
barptitle="Polytomy Counts"

for(patient in unique(polyres$patient)){
# for(patient in c("MD7180","MD7181","MD7182")){
# for(patient in c("MD7180_MPP","MD7181_MPP","MD7182_MPP")){
# for(patient in c("MD7180_LSK","MD7181_LSK","MD7182_LSK")){
 
   tree=PDD[[patient]]$pdx$tree_ml
    tree$edge.length=tree$el.snv
  tree=drop.tip(tree,"zeros")
  plot_tree(get_tree_top(tree,12),cex.label=0);title(patient)

  counts=data.frame(pn=2:max(polytomies$polytomy_number),count=0)
  thispoly=polytomies[which(polytomies$patient==patient),]
  counts$count[match(thispoly$polytomy_number,counts$pn)]=thispoly$count
  #
  barplot(counts$count,names.arg = counts$pn,xlim=c(0,16),las=1,horiz = TRUE,col="black",main=barptitle)
  barptitle=""
  abline(lwd=0.8,col="lightgrey",lty="dashed")
}
```

