---
title: "Tree complexity (shannon diversity)"
author: "Chiraag Kapadia"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE)
knitr::opts_knit$set(root.dir = './scripts/')
```

### Summary

Calculate shannon diversity index for mouse trees and human HSPC trees from Mitchell et al.
Human trees are copied from published code associated with Mitchell et al., available at: https://github.com/emily-mitchell/normal_haematopoiesis/tree/main/8_driver_modelling/input

Ages are normalized accordingly to median lifespan, derived as described in Supplementary Note 1. 

#### Library and support script dependencies

```{r message=FALSE, warning=FALSE}
library(parallel)
library(dplyr)
library(ggplot2)
library(cowplot)
library(phylobase)
library(phangorn)
source("load_and_annotate_tree.R")
source("trees.R")
source("plotting_utils.R")
```

#### Import tree objects for analysis

```{r}
tree_list = readRDS("~/Desktop/somatic_mouse/tree_data/all_tree_list.rds")
```

#### Support function for calculating shannon diversity.
```{r}
shannonDiversityFromTree = function(PD2="",resdir,cutoff,tree=NULL,seed=2024){
  set.seed(seed)
  if(is.null(tree)){
  tree=PD2$pdx$tree_ml
  }
  #downsample to correct for different sized trees (ie normalize 400 colonies vs. 150 colony trees)
  tree=as(subset(as(tree,"phylo4"),tips.include=sample(tree$tip.label,50)),"phylo") 
  tree<- drop.tip(tree,"zeros") #drop zero out branch.
  
  
  whichNodesBelowCutoff = nodeHeights(tree)[,1]>cutoff
  nodes=unique(tree$edge[,1][whichNodesBelowCutoff])
  pp=Descendants(tree,nodes,"tips")  
  uniquesizes=unique(sort(unlist(lapply(pp,length)),decreasing = T))
  if(length(nodes)==0){uniquesizes=c(1)}
  # Ensure that colonies are only calculated based on represented in the largest clade beyond cutoff.
  # ie that a colony is not assessed as part for >1 clade. 
  # This is required for the shannon diversity formula.
  seen=c()
  sizer=c()
  for(i in max(uniquesizes):1){
    if(length(nodes)==0){next}
    for(k in 1:length(pp)){
      if(length(pp[k])<i){
      }else{
        colonies=tree$tip.label[unlist(pp[k])]
        
        if(length(intersect(colonies,seen))<1){
          seen=c(seen,colonies)
          sizer=c(sizer,length(colonies))
        }else{
        }}
    }  
  }
  
  setdiff(tree$tip.label,seen)
  fillin=rep(1,length(  setdiff(tree$tip.label,seen) ))
  counts = c(sizer,fillin)
  proportions= counts/ length(tree$tip.label)
  
  shannon_diversity_index = -1*sum(unlist(lapply(proportions, function(x){x*log(x)})))
  shannon_diversity_index
}
```

#### Calculate shannon diversity for all trees
```{r warning=FALSE}
#Initialize necessary variables 
samples_hsc <- c("MD7180","MD7181","MD7182","MD7183","MD7184","MD7185")
samples_mpp <- c("MD7181_MPP", "MD7182_MPP", "MD7184_MPP", "MD7185_MPP","MD7180_MPP","MD7183_MPP")
samples=c(samples_hsc,samples_mpp)
PDD=tree_list[samples]

human50lifespan=(83.5) #years, see Supp Note 1
mouse50lifespan=28. #months, see Supp Note 1
cutoff=50
seed=1

#Mouse
shannon <- lapply(PDD,FUN=shannonDiversityFromTree,resdir=resdir,cutoff=cutoff,seed=seed)

#Human, for comparison
treefiles=list.files(path="../data/mitchell_trees/",full.names = T)
ages=c(63,29,38,81,77,75,76,48) #years
humans_tree_shannon<-lapply(treefiles,function(x){
  tr=read.tree(x)
  shannonDiversityFromTree(resdir=resdir,cutoff=cutoff,tree = tr,seed=seed)
})


```


#### Plot

```{r}
#Wrangle into appropriate data frame
df_pd<-data.frame(sample=samples,shannon=unlist(shannon))
df_pd$age=-1;df_pd$compartment="";
df_pd$age[which(grepl("80|81|82",df_pd$sample))]<-30
df_pd$age[which(grepl("83|84|85",df_pd$sample))]<-3
df_pd$compartment[which(grepl("MPP",df_pd$sample))]<-"MPP"
df_pd$compartment[which(!grepl("MPP",df_pd$sample))]<-"HSC"
df_pd$PDID<- substr(df_pd$sample,1,6)

df_pd_human_shannon<-data.frame(sample=str_split(treefiles,pattern="tree|_",simplify = T)[,5],#Mind the path subset.
                                shannon=unlist(humans_tree_shannon),
                                age=ages)

#Normalize to lifespan
df_pd$age = df_pd$age/mouse50lifespan;  
df_pd_human_shannon$age=df_pd_human_shannon$age/human50lifespan 

#Now plot
 ggplot(bind_rows(df_pd_human_shannon,df_pd,), aes(x=age,y=shannon,group=compartment))+
  geom_point(aes(color = sample), size=3,position = position_dodge(width=0), stat = "identity")+
  ylab("Shannon Diversity")+
  guides(color=guide_legend(title=""))+
  scale_color_manual(values = myColors)+
  scale_x_continuous(breaks = seq(0,1,.2))+
  scale_y_continuous(breaks = seq(2.6, 4, len = 8),limits = c(2.6,4))+
  xlab("Normalized Lifespan (Percentile)")+
  labs(colour = "Sample")+
  theme_minimal_hgrid()+
  theme(legend.position = "none")

```



